\Title

### 2022-10-11 Grimaldi-etal-BiolCybernetics.tex
Update on Overleaf.

Neuroscience could provide revolutionary answers to these two closely related problems. Indeed, the human brain has the remarkable property of being able to react at any time, while consuming a reasonable amount of energy, about tens of watts. This system is the result of millions of years of natural selection, and a striking difference between biological neural networks and artificial ones is the representation they use. CNNs, for example, represent information passing from one layer to another as tensors representing visual information densely over the visual topography, while representing different properties in different channels. These networks have been known to mimic many properties of biological systems~\citep{schrimpf_brain-score_2020}, yet, this score does not count rapidity or energy. On the other hand, information in a large majority of biological neural networks is represented as spikes, i.e. prototypical all-or-nothing events whose only parameters are their timing and the address of the neuron that emitted this spike~\citep{paugam-moisy_computing_2012}. Why spikes? It is not yet known why most biological systems have evolved to use this representation, but one major advantage seems to be some form of efficiency. In the human brain, each of the approximately 90 trillion nerve cells emits on average one spike per second which fan out to an average of 10000 synapses. Importantly, each neuron fires with great variability, and up to 300 spikes per second. Many attempts have been made to build artificial spiking neural networks, and in particular to achieve the same performance as observed in classical CNNs with a lower energy budget. However, none of these networks has yet surpassed the state of the art performances. Here, we will try to focus on one core property of spiking neurons inspired by neurobiology to progress in this direction.

@@

Let us first review some existing spiking neural network models. A first motivation for these models was to understand biological observations at the different scales of the central nervous system. For example, the Hodgkin-Huxley model describes the dynamics of the membrane potential of a single nerve cell and specifically explains the role of the different voltage-dependent ion channels. It also explains how the model can produce a spike. Larger scale models introduce an simpler dynamics and an explicit spiking mechanism, for example in the Integrate-and-Fire model. It predicts the emergent states of a recurrently connected neural network~\citep{brunel_phase_2000}. Several models have increased the scale of the system to include more neurons, such as The Virtual Brain (TVB)~\citep{sanz_leon_virtual_2013}, which mixes different scales and includes spiking neural networks with (analog) neural mass models. However, these methods are mainly descriptive with applications in neuroscience but few spiking neural models have had an application in computer vision. Another class of approaches have designed normative models of spiking neural networks that aim to have applications in machine learning. One of these is the SpikeNet algorithm which uses a purely temporal approach by encoding information using one spike per neuron~\citep{delorme_spikenet_1999}. Another type of spiking neural network using precise spike timing attempts to determine the structure of the network in order to minimize a cost function. This was implemented in the SpikeProp algorithm~\citep{bohte_error-backpropagation_2002} and has been extanded in novel gradient-based methods. The surrogate gradient method is now widely used in methods that attempt to transfer performance from CNNs to SNNs~\citep{zenke_remarkable_2021}. However, the performance of SNNs is still lagging that of analog networks and the question of the advantage of using spikes in machine learning and computer vision remains open.

@@

This paper explores the possibility that prototypical spiking motifs may be a useful representation used in neurobiology and that we may transfer this hypothesis to neuromorphic algorithms. Our spiking neural network model starts as a simple layer of leaky integrate-and-fire neurons whose connections are defined by a matrix of synaptic weights. Here, we propose to extend our model to a layer of spiking neurons which include heterogeneous delays in addition to weights. In particular, one afferent may be connected with multiple delays. Crucially, we will explicitly use the delay in the computational process. As a consequence, this extends our previous MLR model by encoding spiking motifs as spatio-temporal patterns. The objective in this model, by including the dimension of temporal delays, is to increase the representational capacities of the classifier. In the perspective of building energy-efficient algorithms, we will also titrate quantitatively the best trade-off between robustness and computation time when increasing the number of these heterogeneous delays. %

@@

%%%-----------------------------------------------------------------
%: fig:motion_task
%%%-----------------------------------------------------------------
\begin{figure}%[h!]%[ht!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/motion_task.pdf}
    \caption{
    {\bf Motion detection task.} {\bf (a)} We use large-scale natural images ($512\times512$) in which an aperture ($128\times128$) extracts a sub-image image in the axis of view (the image on the left) such as to reproduce the effect of displacing the eye. To mimic the effect of a saccadic-like eye movement, the axis of view moves following a step-like random walk, and we show here on example path of the trajectory as a green line ($200$ time steps). {\bf (b)} Recording the dynamics of the sub-image as a function of time, it generates a naturalistic movie which may be transformed to an event-based representation. Mimicking the retina, this representation codes for proportional increments or decrements of the luminance in the image, respectively ON (in red) and OFF (in blue) events. This will constitute the input to the spiking neural network.}
    \label{fig:motion_task}
\end{figure}
%%%-----------------------------------------------------------------
Once these eye movement trajectories are generated, we can apply them to a visual scene. For this purpose, we selected a database of 10 natural grayscale images that are commonly used to study the statistics of natural images~\citep{olshausen_emergence_1996}. Note that these are pre-processed to equalize the energy in each frequency band (i.e., whitened). This process is known to occur as early as the retino-thalamic pathway~\citep{dan_efficient_1996}. These images are $512 \times 512$ in size and we will extract sub-images of size $128 \times 128$ which will be positioned around the center of the gaze at each time step (\seeFig{motion_task}-(b)). We will discretize the time in $1~\ms$ bins and produce movies of duration N-T = $250~\ms$. To avoid border effects, we will position a draw of the complete trajectory at random in the image space so that the sub-image is translated using the position given by the trajectory at each time step. The translation is computed using a coordinate roll in the horizontal and vertical dimensions, followed by a sub-pixel translation defined in Fourier space~\citep{perrinet_sparse_2015}. Note that the magnitude of the displacement is relative to the time bin, and we have defined the velocity such that a velocity of $V=1$ corresponds to a motion of one pixel per frame (i.e., per time bin).

@@

%
\section{Discussion}
%%%-----------------------------------------------------------------
%%%-----------------------------------------------------------------
In this paper, we have introduced a generic SNN using heterogeneous delays and have shown how it compares favorably for a visual motion detection task with a state-of-the-art event-based algorithm used for classification. The learned model bears many similarities with neurobiological anatomical observations but also with the results from behavioral results. \note{to change if we compare to ghosh instead of HOTs if we follow that route} The event-driven computations of our method can be reduced drastically through the pruning of synapses, while maintaining top performance for classification. This shows that we may use the precise timing of spikes to enhance neural computations. 
%
\subsection{Synthesis and main contributions}
%%%-----------------------------------------------------------------
To recap, we have introduced in this paper a spiking neural network model with heterogeneous delays that we have trained and evaluated on a complex motion detection task. The model was defined to optimally detect event-driven spatio-temporal motifs. We have shown that when the model is trained on a dataset of natural images with realistic eye movements, the model learns kernels similar to those found in the early visual cortex of humans for example. We have then shown that the model displays similarities with the responses observed in biology. We have evaluated the computational cost of this model if implemented on neuromorphic hardware, showing that the use of heterogeneous delays may be a frugal computational solution for future neuromorphic hardware, but also a key to understand why spikes represent a universal component of neuronal information processing.

>>>>

Neuroscience could provide revolutionary answers to these two closely related problems. Indeed, the human brain has the remarkable property of being able to react at any time, while consuming a reasonable amount of energy, about tens of watts. This system is the result of millions of years of natural selection, and a striking difference between biological neural networks and artificial ones is the representation they use. CNNs, for example, represent information passing from one layer to another as tensors representing visual information densely over the visual topography, while representing different properties in different channels. These networks have been known to mimic many properties of biological systems~\citep{schrimpf_brain-score_2020}, yet, this score does not count rapidity or energy. On the other hand, information in a large majority of biological neural networks is represented as spikes, i.e. prototypical all-or-nothing events whose only parameters are their timing and the address of the neuron that emitted this spike~\citep{paugam-moisy_computing_2012}. Why spikes? It is not yet known why most biological systems have evolved to use this representation, but one major advantage seems to be some form of efficiency. In the human brain, each of the approximately 90 trillion nerve cells emits on average one spike per second which fan out to an average of 10000 synapses. Importantly, each neuron fires with great variability, and up to 300 spikes per second. Many attempts have been made to build artificial spiking neural networks (SNNs), and in particular to achieve the same performance as observed in classical CNNs with a lower energy budget. However, none of these networks has yet surpassed the state of the art performances. Here, we will try to focus on one core property of spiking neurons inspired by neurobiology to progress in this direction.

@@

Let us first review some existing SNN models. A first motivation for these models was to understand biological observations at the different scales of the central nervous system. For example, the Hodgkin-Huxley model describes the dynamics of the membrane potential of a single nerve cell and specifically explains the role of the different voltage-dependent ion channels. It also explains how the model can produce a spike. Larger scale models introduce an simpler dynamics and an explicit spiking mechanism, for example in the Integrate-and-Fire model. It predicts the emergent states of a recurrently connected neural network~\citep{brunel_phase_2000}. Several models have increased the scale of the system to include more neurons, such as The Virtual Brain (TVB)~\citep{sanz_leon_virtual_2013}, which mixes different scales and includes SNNs with (analog) neural mass models. However, these methods are mainly descriptive with applications in neuroscience but few spiking neural models have had an application in computer vision. Another class of approaches have designed normative models of SNNs that aim to have applications in machine learning. One of these is the SpikeNet algorithm which uses a purely temporal approach by encoding information using one spike per neuron~\citep{delorme_spikenet_1999}. Another type of SNN using precise spike timing attempts to determine the structure of the network in order to minimize a cost function. This was implemented in the SpikeProp algorithm~\citep{bohte_error-backpropagation_2002} and has been extanded in novel gradient-based methods. The surrogate gradient method is now widely used in methods that attempt to transfer performance from CNNs to SNNs~\citep{zenke_remarkable_2021}. However, the performance of SNNs is still lagging that of analog networks and the question of the advantage of using spikes in machine learning and computer vision remains open.

@@

This paper explores the possibility that prototypical spiking motifs may be a useful representation used in neurobiology and that we may transfer this hypothesis to neuromorphic algorithms. Our SNN model starts as a simple layer of leaky integrate-and-fire neurons whose connections are defined by a matrix of synaptic weights. Here, we propose to extend our model to a layer of spiking neurons which include heterogeneous delays in addition to weights. In particular, one afferent may be connected with multiple delays. Crucially, we will explicitly use the delay in the computational process. As a consequence, this extends our previous MLR model by encoding spiking motifs as spatio-temporal patterns. The objective in this model, by including the dimension of temporal delays, is to increase the representational capacities of the classifier. In the perspective of building energy-efficient algorithms, we will also titrate quantitatively the best trade-off between robustness and computation time when increasing the number of these heterogeneous delays. %

@@

%%%-----------------------------------------------------------------
%: fig:motion_task
%%%-----------------------------------------------------------------
\begin{figure}%[h!]%[ht!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/motion_task.pdf}
    \caption{
    {\bf Motion detection task.} {\bf (a)} We use large-scale natural images ($512\times512$) in which an aperture ($128\times128$) extracts a sub-image image in the axis of view (the image on the left) such as to reproduce the effect of displacing the eye. To mimic the effect of a saccadic-like eye movement, the axis of view moves following a step-like random walk, and we show here on example path of the trajectory as a green line ($200$ time steps). {\bf (b)} Recording the dynamics of the sub-image as a function of time, it generates a naturalistic movie which may be transformed to an event-based representation. Mimicking the retina, this representation codes for proportional increments or decrements of the luminance in the image, respectively ON (in red) and OFF (in blue) events. This will constitute the input to the SNN.}
    \label{fig:motion_task}
\end{figure}
%%%-----------------------------------------------------------------
Once these eye movement trajectories are generated, we can apply them to a visual scene. For this purpose, we selected a database of 10 natural grayscale images that are commonly used to study the statistics of natural images~\citep{olshausen_emergence_1996}. Note that these are pre-processed to equalize the energy in each frequency band (i.e., whitened). This process is known to occur as early as the retino-thalamic pathway~\citep{dan_efficient_1996}. These images are $512 \times 512$ in size and we will extract sub-images of size $128 \times 128$ which will be positioned around the center of the gaze at each time step (\seeFig{motion_task}-(b)). We will discretize the time in $1~\ms$ bins and produce movies of duration N-T = $250~\ms$. To avoid border effects, we will position a draw of the complete trajectory at random in the image space so that the sub-image is translated using the position given by the trajectory at each time step. The translation is computed using a coordinate roll in the horizontal and vertical dimensions, followed by a sub-pixel translation defined in Fourier space~\citep{perrinet_sparse_2015}. Note that the magnitude of the displacement is relative to the time bin, and we have defined the velocity such that a velocity of $V=1$ corresponds to a motion of one pixel per frame (i.e., per time bin).

@@

%
\section{Discussion}
%%%-----------------------------------------------------------------
%%%-----------------------------------------------------------------
In this paper, we have introduced a generic SNN using heterogeneous delays and have shown how it compares favorably for a visual motion detection task with a state-of-the-art event-based algorithm used for classification. The learned model bears many similarities with neurobiological anatomical observations but also with the results from behavioral results. \note{to change if we compare to ghosh instead of HOTs if we follow that route} The event-driven computations of our method can be reduced drastically through the pruning of synapses, while maintaining top performance for classification. This shows that we may use the precise timing of spikes to enhance neural computations. 
%
\subsection{Synthesis and main contributions}
%%%-----------------------------------------------------------------
To recap, we have introduced in this paper a SNN model with heterogeneous delays that we have trained and evaluated on a complex motion detection task. The model was defined to optimally detect event-driven spatio-temporal motifs. We have shown that when the model is trained on a dataset of natural images with realistic eye movements, the model learns kernels similar to those found in the early visual cortex of humans for example. We have then shown that the model displays similarities with the responses observed in biology. We have evaluated the computational cost of this model if implemented on neuromorphic hardware, showing that the use of heterogeneous delays may be a frugal computational solution for future neuromorphic hardware, but also a key to understand why spikes represent a universal component of neuronal information processing.
